Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-08 Manuscript_Nater_etal.qmd
Added more detailed bullet point structure for discussion

-   Biological results (interpretations and comparisons with other case studies / species / systems)

@@

-   With some focus on what we are able to get insights on thanks to the integration across time and space (i.e. better estimates through data sharing, relative importance of different kinds of variation, etc. ) <!-- @ERLEND & @JAMES, once the "new" results are in: thoughts on what we see that is interesting, cool, expected, unexpected etc. Can we compare to other large scale studies, either on other grouse-like creatures or harvested species more generally? -->

## Implications for management
<!-- preliminary section title) -->

@@

@@

-   Limited usefulness of auxiliary radio-telemetry data as only from one area (big potential for including additional datasets from other small-scale studies, telemetry but also other, e.g. nest monitoring?) <!-- @ERLEND: Any thoughts on what the status here is? Might there be other data available from beyond Lierne? Are there exciting future prospects e.g. linked to automated monitoring with cameras at a larger scale? -->

@@

>>>>

<!-- @ERLEND & @JAMES, once the "new" results are in: thoughts on what we see that is interesting, cool, expected, unexpected etc. Can we compare to other large scale studies, either on other grouse-like creatures or harvested species more generally? -->

<!-- Spatial perspective -->

-   2-55 birds/km^2^, lowest in North, highest in eastern central Norway (close to Swedish border)

-   Most uncertainty in density estimates in areas with relatively low population sizes, low number of transects, and recent start of data collection

-   The model also does "hindcasts", i.e. predictions for years prior to data collection, for areas where data collection started after 2007. But uncertainty increases quickly, so no point looking \> 2 years back.

-   Over half of areas have population increases in recent years.

-   Recent declines: mountains in central Norway, typically following increases up to around 2016

-   Across populations, no clear relationship between density and growth rates (within populations?)

-   There is a spatial association lower density / lower reproduction / higher survival ("slower life history"?) versus higher density / higher reproduction / lower survival ("faster life history?")

-   Survival recruitment trade-off: high survival and low recruitment in far north and southern mountains

<!-- Temporal perspective -->

-   There are some years with relatively high densities in most areas: 2011, 2013, 2018

-   These same years are also characterized by high recruitment, and subsequent low survival, and the match up with the years with big rodent peaks in many areas

-   Many rodents -\> predator release -\> higher recruitment -\> higher densities -\> negative density feedbacks on survival -\> lower survival -\> low density next year?

-   Also: many rodents -\> good year for predators -\> high predation in following season/year -\> lower survival -\> low density next year?

-   This seems to support that the rodent abundance is an important driver -\> supported by the model: rodent effects estimated positive or inconclusive across areas

-   This is well established (for Norwegian ptarmigan and elsewhere?) and was expected

-   BUT: the model was unable to recover the positive relationship between rodent abundance and recruitment when applied only to data from a single location. But with this multi-area model, estimation and inference is much improved -\> space-for-time substitution works!

-   This also means we have potential for including other covariates

-   Examples of what else could matter and why

<!-- Variation beyond rodents -->

-   Including more and/or other covariates may also help further explaining vital rate variation, not least because for both survival and recruitment, residual variation was the largest component.

-   However, looking at spatial vs. temporal variation, we found very different patterns: for recruitment, we have more spatial than temporal variation, but they are not vastly different. For survival, spatial variation was quite low compared to the rather large temporal variation

-   Some hypotheses about large- vs. local-scale drivers?

@@

## Implications for management

<!-- preliminary section title) -->

@@

-   Management also includes optimization of monitoring

-   Detection was less variable across time and space than vital rates, but we do see it tends to be higher in the southern half of the country

-   Discuss this related to landscape features and/or observer pool? Thoughts on what could be done to improve, potentially?

@@

@@

-   Limited usefulness of auxiliary radio-telemetry data as only from one area (big potential for including additional datasets from other small-scale studies, telemetry but also other, e.g. nest monitoring?) <!-- @ERLEND: Any thoughts on what the status here is? Might there be other data available from beyond Lierne? Are there exciting future prospects e.g. linked to automated monitoring with cameras/sound boxes at a larger scale? -->
